---
title: "Relationship between Gross Domestic Product per Capita and Carbon Dioxide Emmissions per Capita among International Countries"
author: "Milo Klein, Shane Egerstrom, Tammy Lee, Brooke Comstock"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
    code-fold: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
code-fold: true
---

# Introduction

Historically, emission producing factories were seen as a symbol of industry and economic progress. Today, many countries who have not historically contributed many CO2 emissions have made the claim in international negotiations that they should be given the opportunity to economically benefit from the burning of fossil fuels before committing to reductions to address climate change. Relating these factors will illuminate how significantly some countries have benefited economically from the burning of fossil fuels. 

## Hypothesized relationship

For the relationship between per capita GDP growth and per capita CO2 emissions, we predict that there will be generally a positive correlation between the two variables. It should be pointed out that we are using CO2 emissions as the response variable and GDP as the explanatory variable.

# Methods

## Data Collection

The CO2 emissions data was taken from the Global Carbon Project, CDAIC and other sources. These sources are used to inform a consumption based projection for historical emissions, that is then related to historical population data. The data documents the change in CO2 emissions produced by each country from 1800-2022, in tonnes of CO2 per capita. (https://docs.google.com/spreadsheets/d/1t_GOXGrXQgSasIB7dWfRrq9dkqjrqwy3xHvywH7ah1A/edit#gid=501532268). 

For the GDP data, this was taken from multiple sources depending on the date range. From 1990 to 2019, data from the World Bank was primarily used, while from before 1990 was taken from Maddison Project Database and Penn World Table of the Groningen Growth and Development Centre. In summary, this data documents the change in GDP per capita for most countries between years 1800 and 2100, in constant U.S. dollars adjusted for 2017 prices. (https://www.gapminder.org/data/documentation/gd001/). 

## Data Cleaning Process and Decisions

After importing both data sets, we cleaned each by making sure all years were numeric and ready to merge. Originally, our CO2 emission data had 194 rows (Countries) and 224 columns (1800:2022). After pivoting longer, there were 43,262 rows (CO2 observations) and 3 columns (country, year, CO2). The data GDP data originally had 195 rows (Countries) and 302 columns (individual years, 1800:2100). After the pivot, there were 58,695 rows (GDP observations) and 3 columns (country, year, GDP). GDP values represented by characters such as “10k” were changed to numeric values. Then, we pivoted both and conducted a full merge by “Year” and “Country”. Finally, we began filtering by years of interest. Future projections (years past 2022) were not considered in our data. 

There were three countries ignored in the calculations: Liechtenstein, San Marino, and Monaco. The datasets only contained data for one metric while missing the other (ex. Liechtenstein only had data for CO2 emissions, but no available data for GDP). In total, 193 countries were evaluated.


```{r, loading}
library(gapminder)
library(gganimate)
library(tidyverse)
library(patchwork)
library(broom)
library(kableExtra)

GDP <- read_csv("gdp_pcap.csv")
Emissions <- read_csv("co2_pcap_cons.csv")
```

```{r, CO2 clean}
Emissions_clean <- Emissions |>
  select(`country`:`2022`)|>
  mutate(across(everything(), ~ str_replace(., "−", "-")))|> 
  filter(across(everything(), ~is.na(.) == 0)) |>
  mutate_at(vars(`1800`:`2022`), as.numeric)|>
  pivot_longer(cols = `1800`:`2022`, 
               names_to = "Year",
               values_to = "CO2")

```

```{r, GDP clean}
GDP_clean <- GDP |>
  select(`country`:`2022`)|>
  mutate(across(`1800`:`2022`, 
                  ~case_when(is.character(.) & str_detect(., "k") ~ as.numeric(str_replace(., "k", "")) * 1000,
                             TRUE ~ as.numeric(.)))) |>
  pivot_longer(cols = `1800`:`2022`,
               names_to ="Year", values_to = "GDP")
```

```{r, merge}
merged_data <- full_join(GDP_clean, 
                         Emissions_clean, 
                         by = c("country", "Year")) |>
               mutate(Year = as.numeric(Year))
```

# Results

## Observed Data and Fitted Linear Regression Model Visualizations

There are multiple ways to model the relationship between GDP and CO2. However, a great factor in the increase of CO2 emissions is the beginning of the Industrial Revolution, with some countries starting as early as the 19th century.

We sought to evaluate the changing relationship between average country GDP per capita and CO2 emissions per capita by fitting our observations with a linear regression model: both for overall averages in the dataset and for incremental averages for each year, starting from before industrialization in 1800, to 2010, the present day.

A linear regression model looks at one explanatory/independent variable and see how it affects the values of one response/dependent variable. A "linear" regression specifically implies that changes in the explanatory variable will cause changes to the dependent variable at a fixed function increment, creating "line." Some linear regression models may also account for the data "noise" to account for random variation or error between the independent and dependent variables, while others may be fitted, where the function would have to estimate values.

```{r fig.height=40, fig.width=5}

#1: Visual exploring the relationship between the two quantitative variables changing over time
   
# Create a new variable for decade
filtered_data_decade <- merged_data |>
  mutate(decade = (10*(Year %/% 10))) |>
  filter(decade %in% 1800:2019) |>
  group_by(country, decade)|>
  summarize(GDPavg=mean(GDP), CO2avg=mean(CO2))

# Create a line plot for the selected countries faceted by decade
filtered_data_decade %>%
  ggplot(aes(x = GDPavg, y = CO2avg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ decade, scales = "free_y", ncol = 1) +  # Stacking each facet
  labs(title = "Trends of GDP per capita and CO2 Emissions per capita",
       x = "GDP (US $)",
       y = "",
       subtitle = "CO2 Emissions [tons per person]")
```

```{r}
gif <- ggplot(filtered_data_decade, aes(GDPavg, CO2avg)) +
  geom_point(color = "steelblue", alpha = 0.7, show.legend = FALSE) +
  geom_smooth(method = "lm", se = FALSE) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}',
       x = 'GDP per Capita (dollars)',
       y = 'CO2 Emissions per Capita (tons)') +
  transition_time(decade) +
  ease_aes('linear') # Tells the animation to progress at a linear rate

animate(
  plot = gif,
  renderer = gifski_renderer("CO2_vs_GDP.gif")
)
```

The relationship between GDP and CO2 per capita remains linear across each decade analyzed (as shown in the plots above). Therefore, it is apparent that for most countries, as the decades progressed and per capita GDP increased, so did the per capita CO2 emissions. This trend became noticeably more pronounced beginning in 1940 (after World War II), when many countries' GDP began increasing rapidly.

```{r, Visual_2}
#2: Visual exploring the relationship between the two quantitative variables you are investigating

# Group by country and calculate average GDP and CO2 emissions
average_data <- merged_data |>
  group_by(country) |>
  summarise(avg_GDP = mean(GDP, na.rm = TRUE),
            avg_CO2_emissions = mean(CO2, 
                                     na.rm = TRUE)
            )|>
  filter(across(everything(), ~is.na(.) == 0))

# Create a scatter plot:
side_by_side_1 <- average_data %>%
  ggplot(aes(x = avg_GDP,
             y = avg_CO2_emissions,
             label = country)
         ) +
  geom_point(color = "red") +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "blue",
              linetype = "dashed") +  # Add linear regression line
  geom_text(nudge_x = 0.1,
            nudge_y = 0.1,
            check_overlap = TRUE) +
  labs(title = "Observed Data",
       subtitle = "Average Relationship Between GDP and CO2 Emissions per capita",
       x = "Average GDP per capita (US $)",
       y = "Average CO2 Emissions [tons per person]")

side_by_side_1
```
Out of the 193 countries, most countries fall in the area with lower average GDPs and lower CO2 emissions; it's also in this area that the "line"/regression model fits more appropriately as the data point entries are close to the line. However, as average GDP of a country grows, not only are there less countries/datapoints, but there are also many more outliers deviating from model. These outliers may either be countries that produce less expected CO2 emissions per unit of GDP (Brunei, Switzerland, Andorra) or countries that produce more expected CO2 emissions per unit of GDP (Luxembourg, Kuwait, USA).


Both visualizations exhibit a positive relationship between average GDP and average CO2 emissions. From the first visualization, there are clear differences in the "slopes" of the models between each decade. The second visualization shows a clear positive linear relationship between countries who historically have the highest GDPs and the countries that have historically emitted the most CO2. 

## Linear Model Regression Interpretation

A linear regression model is appropriate for this analysis because we assumed that the residuals of the data are normally distributed. Based on our histogram of R^2 distributions below, this interpretation holds true. Furthermore, we selected a linear model because it provided a simple way of conveying the strength and direction of the independent/dependent variables. 

```{r, Regression_Data_V2}
# Regression Data for Visual #2:

# Fit a simple linear regression model
regression_model <- lm(avg_CO2_emissions ~ avg_GDP, data = average_data)

# Display the summary of the regression model
#summary(regression_model)
broom::tidy(regression_model) |>
  kbl() |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

A general fitted linear regression model, estimating deviations from noise, may be represented as:

$$  
\Large 
 \hat{y} = \beta_0 + \beta_1 x
$$

where $\hat{y}$ is the response variable, $\beta_0$ is the y-intercept of the linear model, $\beta_1$ is the slope of the linear model, and x is the explanatory variable.

From the results from the coefficient table, the estimated regression model between average GDP per capita and average CO2 emissions per capita can be represented as:

$$  
\Large 
 \hat{y} = -0.2863 +  0.0003x
$$
The estimated slope coefficient for average GDP is 0.0003, indicating that, on average, a one-unit increase in average GDP is associated with a y-unit increase/decrease in average CO2 emissions; tangibly, that would mean for every $1000 increase in average GDP/capita, on average, we would expect to see an increase of 0.3342 tons of CO2/capita. The y-intercept is -0.2863.

The p-value for the average GDP coefficient is less than 0.05, indicating that average GDP is a statistically significant predictor of average CO2 emissions.

We can assess how “good” a regression model is by looking for how much of the variability in the response values were accounted for in the regression model, specifically looking at the regression model's fitted values. Meanwhile, the variance in the residual values represent the amount of variability not accounted for in the model.

```{r}
variance_table <- data.frame(
  Component = c("Variance in the response values", "Variance in the fitted values", "Variance in the residuals"),
  Variance = c(var(regression_model$model$avg_CO2_emissions), 
               var(regression_model$fitted.values),
               var(regression_model$residuals)
               )
  )
  
variance_table |>
  kbl() |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The variance of the fitted values accounts for nearly 76.5% of variance response values while the residual variance accounts for 23.9%. This signifies that the linear regression model is a good fit for our observations, explaining that 76.5% of variability in average CO2 emissions is explained by the model (matching with the outputted R-squared value of 0.76).

## Simluated Results from Estimated Linear Regression Model
The effectiveness of our linear regression model can be further gauged by seeing the differences between the observed and simulated data created from the linear regression model. We try to replicate noise found in the observed data by adding random errors to the prediction.

```{r, Section_3-1, fig.width=14}
set.seed(1984)

average_data <- merged_data |>
  group_by(country) |>
  summarise(avg_GDP = mean(GDP, na.rm = TRUE),
            avg_CO2_emissions = mean(CO2, 
                                     na.rm = TRUE)
            )|>
  filter(across(everything(), ~is.na(.) == 0))

noise<-function(x, sd){
  x + rnorm(length(x),mean=0,sd)
}

CO2_predict<-predict(regression_model)
CO2_sigma<-sigma(regression_model)
model_w_noise<-noise(CO2_predict, CO2_sigma)

simulation<- tibble(sim_CO2 = model_w_noise)

pred_obs_df<-average_data|>
  mutate(pred_CO2=simulation$sim_CO2) 

side_by_side_2 = pred_obs_df |>
  ggplot(aes(x = avg_GDP, 
             y = pred_CO2, 
             label = country)
         ) +
  geom_point(color = "orange") +
  geom_smooth(method = "lm", 
              se = FALSE, 
              color = "blue", 
              linetype = "dashed") +
  geom_text(nudge_x = 0.1, 
            nudge_y = 0.1, 
            check_overlap = TRUE) +
  labs(title = "Simulation",
       x = "Average GDP per capita (US $)",
       y = "Predicted CO2 Emissions [tons per person]")

# Arrange plots side by side
side_by_side_1 + side_by_side_2 + plot_layout(ncol = 2)

```

Based on the plots shown above, one instance of simulated data predicts fewer outliers at larger (above approximately ~$15000) average GDP values, congregating them closer to the modeled slope. In addition, the simulation did output negative values for a couple of the lower GDP/capita countries originally recorded in the observed. Overall though, the two plots are similar in that both predict a positive-linear trend between average per capita CO2 emissions and GDP, with more data points near the lower GDPs and more variation at the higher GDPs.

## One Thousand Simulations
To further challenge the fit of the linear regression model, instead of just one instance, one-thousand linear models from the simulated data and random noise can be run.

```{r, Section_3-2}
set.seed(1984)
sims <- map_dfc(0:1000, #if you want to practice run, don't use 1000.................. :(
                ~(noise(x = CO2_predict , sd = CO2_sigma))) #runs through noise() 1000 times, rnorm() creates different values

colnames(sims) <- colnames(sims) |> #changes column names to something more readable
  str_replace(pattern = "\\.\\.\\.",
                  replace = "_")

thousand_sims <- average_data |>  #combines the 1000 simulations with the original avg CO2 values, its technically 1000 sims + 1 obs
  filter(!is.na(avg_GDP), 
         !is.na(avg_CO2_emissions)) |> 
  select(avg_CO2_emissions) |> 
  bind_cols(sims)

```

```{r}
# Plot of R^2:
thousand_r2s <- thousand_sims |> # runs the lm() method 1000 times and extracts the R2 values into new... matrix???
  map(~ lm(avg_CO2_emissions ~ .x, data = thousand_sims)) |> 
  map(glance)|>
  map_dbl(~ .x$r.squared)

thousand_r2s <- thousand_r2s[names(thousand_r2s) != "avg_CO2_emissions"] #get rid of the "avg_CO2_emissions" column

tibble(sims = thousand_r2s) |> #plot R2 value histogram
  ggplot(aes(x = sims)) + 
  geom_histogram(binwidth = 0.025, fill = "salmon", alpha = 0.8) +
  labs(x = expression("Simulated"~ R^2),
       y = "",
       subtitle = "Number of Simulated Models, Total (n = 1000)",
       title = "Distribution of Simulated Linear Model\nR-Squared Values of Country GDP per capita to CO2 emissions per capita") +
  theme_bw() +
  scale_x_continuous(breaks = seq(0.35, 0.75, 0.025))
```

As shown in the plot above, the simulation results are centered around an R-squared value of between 0.575 and 0.600. This indicates that the model fits the data fairly well, because the independent variable (GDP per capita) can explain 57.5% to 60% of the variability observed in the dependent variable (CO2 emissions per capita).


