---
title: "Project_Proposal_Group3"
author: "Milo Klein, Shane Egerstrom, Tammy Lee, Brooke Comstock"
format: html
editor: visual
---

## Hypothesized relationship (and any outside references):
We predict that there will be generally a positive correlation between GDP growth and CO2 emissions.

## Detailed data and variable description:
* CO2 data: The CO2 emissions data was taken from the Global Carbon Project, CDAIC and other sources. It documents the change in CO2 emissions produced by each country from 1800-2022. (https://docs.google.com/spreadsheets/d/1t_GOXGrXQgSasIB7dWfRrq9dkqjrqwy3xHvywH7ah1A/edit#gid=501532268)

  + Originally had 195 rows (Countries) and 302 columns(1800:2100)
  + After pivot: 58,695 rows (GDP observations) and 3 columns (country, year, GDP)

* GDP data: This data was taken from multiple sources, depending on the date range. From 1990 to 2019, data from the World Bank was primarily used, while from before 1990 was taken from Maddison Project Database and Penn World Table of the Groningen Growth and Development Centre. In summary, this data documents the change in GDP for most countries\
between years 1800 and 2100. (https://www.gapminder.org/data/geo/changes/)
  + Originally had 194 rows (Countries) and 224 columns (1800:2022)
  + After pivot: 43,262 rows (CO2 observations) and 3 columns (country, year, CO2)
  
## Data cleaning process and decisions:
After importing both data sets, we cleaned each by making sure all years were numeric and ready to merge. GDP values represented by characters such as “10k” were changed to numeric values. 

We also eliminated an “X” character from each of the years in the Emissions data set that prevented merging.

We then pivoted both and conducted a full merge by “Year” and “Country”.

Also began filtering by years of interest. Future projections (years past 2021) were not considered in our data.

```{r, loading}
library(tidyverse)
GDP <- read_csv("gdp_pcap.csv")
Emissions <- read_csv("co2_pcap_cons.csv")
```

```{r, CO2 clean}
Emissions_clean <- Emissions |>
  rename_with(~gsub("^X", "", .), everything()) |>
  mutate_at(vars(`1800`:`2022`), as.numeric)

Emissions_longer <- pivot_longer(Emissions_clean, cols = `1800`:`2022`, 
                                  names_to = "Year", values_to = "CO2")


```

```{r, GDP clean}
country_names<-GDP[1]

clean_gdp_value <- function(df) {
  df |>
    mutate(across(everything(), ~case_when(is.character(.) & str_detect(., "k") ~ as.numeric(str_replace(., "k", "")) * 1000,
                            TRUE ~ as.numeric(.))))
}

GDP_clean <- clean_gdp_value(GDP)
GDP_clean[1] <- country_names

GDP_longer <- pivot_longer(GDP_clean, cols = `1800`:`2100`,
                            names_to ="Year", values_to = "GDP")

```

```{r, merge}
merged_data <- full_join(GDP_longer, Emissions_longer, by = c("Year", "country"))
```

